---
title: "Riskmetric survey package selection"
author: "Eric Milliman"
date: "1/19/2022"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 2
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(gt)
library(tidyverse)
library(ctv)
library(igraph)

graph.reverse <- function (graph) {
  if (!is.directed(graph))
    return(graph)
  e <- get.data.frame(graph, what="edges")
  ## swap "from" & "to"
  neworder <- 1:length(e)
  neworder[1:2] <- c(2,1)
  e <- e[neworder]
  names(e) <- names(e)[neworder]
  graph.data.frame(e, vertices = get.data.frame(graph, what="vertices"))
}

ctvs <- available.views(repos = "https://cran.rstudio.com/")
views <- sapply(ctvs, function(x) x$name)[c(1,3,5,7,18,21,23,28,36,38)]

ctv_pkgs <- bind_rows(lapply(ctvs[c(1,3,5,7,18,21,23,28,36,38)], 
                             function(x) x$packagelist))
ctv_pkgs <- ctv_pkgs$name[!ctv_pkgs$core]

if(!file.exists("ap.rds")){
  ap <- as.data.frame(available.packages(repos = "https://cran.rstudio.com/"))
  saveRDS(ap, "ap.rds")
} else {
  ap <- readRDS("ap.rds")
}

if(file.exists("downloads_1_yr_from.rds")){
  dwlds <- readRDS("downloads_1_yr_from.rds")
} else{
  dwlds <- lapply(split(ap$Package, ceiling(seq_along(ap$Package)/500)), 
                  function(x){
                    cranlogs::cran_downloads(x, from = Sys.Date()-366, 
                                             to = Sys.Date()-1)
                  })
  
  dwlds <- do.call(rbind, dwlds)
  saveRDS(dwlds, "downloads_1_yr_from.rds")
}
```

# Purpose

The __riskmetric__ package aims to provide an objective assessment of package quality and risk. During the development of this package we recognized that there is an opportunity to empirically learn which package assessments are most important to the community with regards to package risk/quality. To gather the necessary data to learn these weights empirically we have designed a simple survey which asks participants to rate the quality/risk a small subset of packages based on their personal experience/opinion.

## Method to choose packages for survey

We cannot gather quality ratings for all packages in R for a number of:
  1) There are more than 15,000 packages in CRAN, thousands more in Bioconductor, and an untold number in github/bitbucket/gitlab. And our survey population is not that large.
  2) Pharma/biotech's biggest focus is statistical analysis and reporting, which many packages do not get used by the industry (e.g. finance, GIS focised packages).
  3) Lastly, a typical end user may only be familliar with the packages they invoke and not their dependencies (e.g. most users are probably familiar with ggplot2, but not rlang, digest, and scales). 

In the below section we will walk through the section of packages to survey users with, keeping the above criteria in mind.

## CRAN views

To narrow down the list of packages to survey users about we selected a handful CRAN Task Views that Pharma/Biotech users most aligh with.

```{r}
vws <- sapply(ctvs, "[[", 1)
gtobj <- bind_cols(split(vws, cut(seq_along(vws), 4, labels = F))) %>% 
  gt() %>% 
  tab_header("CRAN Task Views") %>%  
  tab_options(column_labels.hidden = TRUE)

for(i in seq_along(gtobj$`_data`)){
  gtobj <- gtobj %>% tab_style(
    style = cell_fill("yellow"),
    locations = cells_body(
      columns = all_of(i),
      rows = gtobj$`_data`[[i]] %in% vws[c(1,3,5,7,18,21,23,28,36,38)]
    )
  )
}

ap <- ap %>% mutate(isStat=+(Package %in% ctv_pkgs))

gtobj
```

## Package popularity

People are more likely to be familiar with more popular packages (i.e. packages with high download counts). So we pulled the download statistics from the past year for all CRAN packages, to use for identifying the most popular packages.
```{r}
ap <- dwlds %>% 
  group_by(package) %>% 
  summarise(count=sum(count)) %>% 
  right_join(., ap, by = c("package" = "Package"))

ap %>% ggplot(aes(x=count)) + 
  geom_histogram() + scale_x_log10() + 
  xlab("Downloads (log10)") +
  ggtitle("All CRAN Packages", "Last 365 days")
```

If we look at the download statistics for the packages in the selected CRAN Task Views we see that the distribution of download counts is similar to all CRAN pacakges

```{r}
ap %>% 
  filter(isStat==1) %>% 
  ggplot(aes(x=count)) + 
  geom_histogram() + scale_x_log10() + 
  xlab("Downloads (log10)") +
  ggtitle("Selected CRAN TaskViews", "Last 385 days")

```

## Packages that are most-likely user-facing

Some packages are popular by virtue of being heavily depended on (e.g. Rcpp, rlang, etc). However, an average user may not be familiar with these packages because they are not directly exposed to the user. To tease apart "low-level" packages from more user-facing packages, we assess the number of reverse imports and reverse dependencies a package has. Packages with alot of reverse Imports suggests it maybe a lower level package that users may not use directly (i.e. `rlang`. Packages with a lot of reverse Depends on the other hand are packages that are more likely to be used directly by the user (i.e. `ggplot2`). Ideally we want pakcages with high downloads, a low number of reverse imports and high number of reverse depends (though this last criteria is less important).

```{r}
dg <- miniCRAN::makeDepGraph(ap$package, suggests = FALSE)
dgr <- graph.reverse(dg)

ap_df <- left_join(ap, 
                data.frame(package = V(dg)$name,
                           revImports=degree(delete_edges(dg, E(dg)[which(E(dg)$type!="Imports")])),
                           revDeps=degree(delete_edges(dg, E(dg)[which(E(dg)$type!="Depends")])),
                           pageRank = page_rank(dgr, directed = TRUE)[[1]]
                           ) 
                )

ap_df %>%
  ggplot(aes(x=revImports+0.1)) +
  geom_histogram() + scale_x_log10() +
  ggtitle("Package Reverse Imports","All CRAN")

ap_df %>%
  ggplot(aes(x=revDeps+0.1)) +
  geom_histogram() + scale_x_log10() +
  ggtitle("Package Reverse Depends","All CRAN")

ap_df %>% 
  ggplot(aes(x=revDeps+0.1, y=revImports+0.1)) + geom_hex() +
  scale_x_log10() + scale_y_log10() + ggtitle("Relationship between Reverse imports and reverse depends", "All CRAN packages")

ap_df %>% 
  ggplot(aes(x=count, y=revImports+0.1)) + geom_hex() +
  scale_x_log10() + scale_y_log10() + ggtitle("Relationship between Reverse imports and downloads", "All CRAN packages")

ap_df %>% 
  ggplot(aes(x=count, y=revDeps+0.1)) + geom_hex() +
  scale_x_log10() + scale_y_log10() + ggtitle("Relationship between reverse depends and downloads", "All CRAN packages")

```

## Packages selected

Using the criteria above we have selected the following packages, with the aim to keep the list to a few hundred.

```{r}
ap_df_filtered <- ap_df %>% 
  mutate(count_rank=percent_rank(count)) %>%
  group_by(isStat) %>% 
  mutate(revDep_rank = percent_rank(revDeps),
         count_rank_g = percent_rank(count)) %>% 
  filter((revDeps>0 & revImports==0 & count_rank >=0.99) | 
           (isStat==1 & revDep_rank >= 0.75 & count_rank_g >=0.75 ) | 
           (isStat==0 & revDep_rank >= 0.98 & count_rank_g >=0.98))


ap_df_filtered %>% 
  select(package, inCTV=isStat, downloads=count, revDeps, revImports) %>% 
  DT::datatable()
saveRDS(ap_df_filtered, "packages_selection_current.rds")
saveRDS(ap_df_filtered, paste0("packages_selection_", gsub("-","", Sys.Date(), ".rds")))
```